Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2012-11-28 large_number_squeezing_with_atoms.tex
The edge of what I can write/claim without results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass{iopart}
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\usepackage{graphicx}
\usepackage{ulem}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{subcaption}

@@

\section{Multimode analysis of squeezing}

@@

A one-dimensional simulation is sufficient to examine the hypothesis that the squeezing will be higher for a BEC where the spatial mode has a more constant density profile.  We compare squeezing for fields starting in the ground states of a harmonic trap with negligible nonlinearity (a Gaussian), a harmonic trap with strong nonlinearity (Thomas-Fermi), and a constant potential (constant density). Using (\ref{eqChiUequivalence}) we find the following equivalences
\begin{eqnarray}
\chi &=& U \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}} \,\,\,\, ({\textrm{Gaussian}}) \\
%
\chi &=& \frac{4}{7} \left( \frac{15 U \omega_x \omega_y \omega_z}{16 \pi \sqrt{2}} \right)^{2/5} \left( \frac{m}{N} \right)^{3/5} \,\,\,\, ({\textrm{Thomas-Fermi}}) \\
%
\chi &=& \frac{U}{V} \,\,\,\, {\textrm{(constant density)}}
\end{eqnarray}
\textbf{Choosing parameters where the mode shape of the multimode system doesn't change much during the evolution allows us to compare the effect of mode shape on the best squeezing that can be attained relative to the two mode system.  Joe: We need to be more specific here.}

The results of this simulation are shown in Figure~\ref{figModalSqueezingEffects1D}, which clearly confirms the hypothesis that . Clearly the closer the mode shape is to constant density, the better.

\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/modal_squeezing_effects_1D.eps}
    \caption{Effects of mode shape on squeezing. The plot shows the normalized number variance in mode $\hat{a}$ as a function of the time the final coupling pulse is applied. Thick red line, analytic two-mode solution; solid blue line, constant density mode; dashed line, Thomas-Fermi mode; dash-dotted line, Gaussian mode. The nonlinearities of the various 1D simulations have been adjusted for their mode shape so that they are equivalent to that of the 0D, two-mode mode nonlinearity. Parameters: $n_a = n_b =1 \times 10^5$, $\chi_{aa}=0.03 \hbar, \chi_{ab}=\chi_{bb}=0, \tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=4.0$. }
    \label{figModalSqueezingEffects1D}
\end{figure}

As it can be difficult to engineer specific mode shapes, another alternative is to post-select a mode. For example, rather than considering the variance in the total number of atoms in a mode, it may be possible to only count, say the atoms in the central portion of the atomic cloud. For example, suppose one had a gaussian atomic distribution, but one considers a mode that only includes the atoms within $p$ standard deviations of the centre. For low $p$, this mode shape is considerably closer to the constant density case than a standard gaussian. The equivalence between the single mode case and this mode is given by
\begin{equation}
\chi = U \frac{{\mathrm{Erf}} (\sqrt{2p})^3}{{\mathrm{Erf}}(p)^6} \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}}
\end{equation}
And example comparing the best attainable squeezing for this mode compared to a single gaussian is shown in Figure~\ref{XXX}.

\section{Using a $\pi$ pulse to improve mode-matching}
\begin{itemize}
  \item Say we can use a pi pulse to swap the populations half way through, so if the states have different nonlinearities we still end up with the same mode shapes when we recombine. The fact that this works will be clear from the SM equations for a(t), b(t) when the beam splitter is on
  \item show some plots of squeezing with and without the pi pulse
\end{itemize}

\section{Squeezing in 2D and 3D}
  \begin{itemize}
  \item Two issues. 1) as we add dimensions results get worse (maybe point out how cool we are for being able to do multimode 3D simulations). 2) As our volume increases results get worse.
  \item To gain insight into this we do a bog analysis
  \item Do a comparision plot between SM analytic model and Bog to show everything agrees
  \item show that the number of mode available degrades squeezing
  \item Plot some bog results v MM sims for increasing box sizes, and increasing dimensions
  \item Can we put some numbers on how bad 3D is compared to 2D, in order to motivate tight trapping?
  \end{itemize}

@@

\section{Mitigating multimode quantum statistical effects}
  \begin{itemize}
  \item One way to approximate constant density is to look at core of BEC, since if we only look at number in the central 1 sigma portion of gaussian it's close to flat; show results compared to single mode equivalent

  \item Another improvement is to put the BEC in a box since then we get constant density, discuss feasibility
  \item 3D still sucks even with box, so point out we can win by freezing out dimensions
  \item Show derivation for what physical situations are needed to freeze out 1, 2 and 3 dimensions. In general we need small number, which makes it hard since we want large number squeezing. Large number requirement leads to implausible geometry in 1D (i.e. a tube with a ridiculous aspect ratio), but going from 3D down to 2D is probably do-able
  \end{itemize}

@@

The criteria we need to accomplish this depends on the number of dimensions we want to freeze out, so investigate the cases where we reduce a 3D system to an effective 0D, 1D, and 2D system seperately. In all cases we assume we have a box potential with length $L$ in the dimension(s) that aren't frozen out, and tight harmonic trapping potential of frequency $\omega$ in the frozen dimensions.

@@

\ack
Acknowledgements in here; funding sources etc

@@

\appendix
\section{Derivation of two mode squeezing formula}
\label{appendixTwoModeDerivation}
In this Appendix we derive the equations for the number variances which were earlier stated without proof. We use the model described in Section \ref{secTwoModeAnalytic}, where we consider a two mode system with $N$ particles, initially prepared as a coherent state with all the population in mode $|a\rangle$ at time $t=0$. A coupling pulse is then applied, so that the system evolves under the Hamiltonian
\begin{equation}
\hat{H_1} = \hbar \omega \hat{a}^{\dagger} \hat{a} +  \hbar \omega \hat{b}^{\dagger} \hat{b} 
          + \Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger} \hat{a} )
\end{equation}
until time $t=t_1$. From time $t_1$ until time $t=t_2$ the coupling is turned off, and the system evolves under the Hamiltonian
\begin{equation}
\hat{H_2} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}.
\end{equation}
Finally, from time $t_2$ until time $t=t_3$, the coupling is once again applied with a phase offset of $\phi$ relative to the first time, and the system evolves via
\begin{equation}
\hat{H_3} = \hbar \omega \hat{a}^{\dagger} \hat{a} +  \hbar \omega \hat{b}^{\dagger} \hat{b}
          + \Omega (e^{i\phi} \hat{a}^{\dagger} \hat{b} + e^{-i\phi} \hat{b}^{\dagger} \hat{a} ).
\end{equation}
It is clear that the linear self-energy terms proportional to $\omega$ result in a constant phase rotation of the fields $\hat{a}$ and $\hat{b}$. As this system is self-contained and global phase is unmeasurable, all phase measurements can only rely on the difference in phase between the two modes. This phase rotation is therefore irrelevent and will be ignored in the following analysis.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass{iopart}
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\usepackage{graphicx}
\usepackage{ulem}
\usepackage{amssymb}
\usepackage{caption}
\captionsetup{compatibility=false}
\usepackage{subcaption}

@@

\section{Multimode analysis of squeezing} \label{sec:MMdescription}

@@

A one-dimensional simulation is sufficient to examine the hypothesis that the squeezing will be higher for a BEC where the spatial mode has a more constant density profile.  We compare squeezing for fields starting in the ground states of a harmonic trap with negligible nonlinearity (a Gaussian), a harmonic trap with strong nonlinearity (Thomas-Fermi), and a constant potential (constant density). Using (\ref{eqChiUequivalence}) we find the following equivalences
\begin{eqnarray}
\chi &=& U \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}} \,\,\,\, ({\textrm{Gaussian}}) \\
%
\chi &=& \frac{4}{7} \left( \frac{15 U \omega_x \omega_y \omega_z}{16 \pi \sqrt{2}} \right)^{2/5} \left( \frac{m}{N} \right)^{3/5} \,\,\,\, ({\textrm{Thomas-Fermi}}) \\
%
\chi &=& \frac{U}{V} \,\,\,\, {\textrm{(constant density)}}
\end{eqnarray}
%
Choosing parameters such that the evolution is sufficiently fast that the mode shape is reasonably stable over the simulation gives the best squeezing, and allows us to make a fair comparison between the squeezing achievable assuming that mode-matching is near optimal.  The results of this simulation are shown in Figure~\ref{figModalSqueezingEffects1D}, which clearly confirms the hypothesis that maximum squeezing is obtained when the mode shape is closest to constant density.  In practice, this isThis is largely due to the fact that the squeezing rates become non-uniform, but slower dynamics will also lead to issues with mode-matching.

\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/modal_squeezing_effects_1D.eps}
    \caption{Effects of mode shape on squeezing. The plot shows the normalized number variance in mode $\hat{a}$ as a function of the time the final coupling pulse is applied. Thick red line, analytic two-mode solution; solid blue line, constant density mode; dashed line, Thomas-Fermi mode; dash-dotted line, Gaussian mode. The nonlinearities of the various 1D simulations have been adjusted for their mode shape so that they are equivalent to that of the 0D, two-mode mode nonlinearity. Parameters: $n_a = n_b =1 \times 10^5$, $\chi_{aa}=0.03 \hbar, \chi_{ab}=\chi_{bb}=0, \tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=4.0$.  Clearly, the maximum squeezing is achieved when the atomic density profile is flattest.}
    \label{figModalSqueezingEffects1D}
\end{figure}

One way to avoid the requirement to engineer and maintain specific atomic modes is to post-select a mode.  Rather than considering the variance in the total number of atoms in a mode, it may be possible to consider only the statistics of a spatially filtered area of the atomic cloud.  For example, suppose the atoms were in a Gaussian atomic distribution, but we consider a mode that only includes the atoms within $p$ standard deviations of the centre. For low $p$, this mode shape is considerably closer to the constant density case than a standard Gaussian. The equivalence between the single mode case and this mode is given by
\begin{equation}
\chi = U \frac{{\mathrm{Erf}} (\sqrt{2p})^3}{{\mathrm{Erf}}(p)^6} \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}}
\end{equation}
A simulations comparing the best attainable squeezing for this mode compared to an unfiltered Gaussian is shown in Figure~\ref{XXX}.  We can see that the squeezing is higher.  One obvious disadvantage to post-selecting modes is that it can severely reduce the effective number of atoms.  This is a serious issue for most applications of squeezed sources, such as interferometry, where the signal to noise scales with the square root of the flux.

\section{Using a $\pi$-pulse to improve mode-matching}
Modal mismatch is generated by the different atomic potentials seen by the two internal states.  We propose a straightforward method of minimising this effect: using a $\pi$-pulse to swap the internal states after half of the evolution.  In the limit where each state stayed in a single mode (although that mode might change), and where the nonlinear terms did not change with time, this method would produce identical mode shapes for both pulses after the hold time.  In practice, neither of these conditions will hold, particularly when the squeezing is completely removed by multimode effects, but the $\pi$-pulse should produce some improvement in regimes where the squeezing is only moderately reduced.

\begin{itemize}
  \item show some plots of squeezing with and without the pi pulse
\end{itemize}

In Fig.~\ref{piPulseFig} we show the effects of a $\pi$-pulse on the squeezing generated \textbf{by some specific system}.

\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/piPulseFig.eps}
    \caption{This is a picture of a $\pi$-pulse working.}
    \label{piPulseFig}
\end{figure}

\section{Squeezing in 2D and 3D}
The simulations in the previous two sections were one dimensional, and show how having spatially constant density and using a $\pi$-pulse will improve the resultant squeezing.  While these conclusions remain valid for simulations in two or three dimensions, adding each additional dimension results in a degradation of the squeezing.  This is true even when the potential (and the resulting atomic density) is constant across the bulk of the trap.  Furthermore, the expected squeezing goes down as the volume of the trap increases.  This can be seen in Fig.~\ref{1D2D3DworsererFig}, where we show the squeezing generated \textbf{by 1D, 2D and 3D BECs in a box with some parameters}.

\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/1D2D3DBox.eps}
    \caption{Even BECs in a box get worse with both dimension and volume.}
    \label{1D2D3DworsererFig}
\end{figure}

This dependence on dimensionality and volume are not explained by our arguments in section \ref{sec:MMdescription}, which focus on the effects of inhomogeneity within the trap.  Fundamentally, this degradation of the squeezing is due to nonlinearity-induced coupling between different spatial modes.  As the dimension and volume increases, the spectral density of available excited modes increases, leading to a reduction in squeezing.

Excitations of a BEC are described by the Bogoliubov modes...

  \begin{itemize}
  \item Introduce Bog analysis
  \item Do a comparision plot between SM analytic model and Bog to show everything agrees
  \item show that the number of mode available degrades squeezing
  \item Plot some bog results v MM sims for increasing box sizes, and increasing dimensions
  \item Can we put some numbers on how bad 3D is compared to 2D, in order to motivate tight trapping?
  \end{itemize}

@@

\section{Mitigating multimode quantum statistical effects}
  \begin{itemize}
  \item 3D still sucks even with box, so point out we can win by freezing out dimensions
  \item Show derivation for what physical situations are needed to freeze out 1, 2 and 3 dimensions. In general we need small number, which makes it hard since we want large number squeezing. Large number requirement leads to implausible geometry in 1D (i.e. a tube with a ridiculous aspect ratio), but going from 3D down to 2D is probably do-able
  \end{itemize}

@@

The criteria we need to accomplish this depends on the number of dimensions we want to freeze out, so investigate the cases where we reduce a 3D system to an effective 0D, 1D, and 2D system separately. In all cases we assume we have a box potential with length $L$ in the dimension(s) that aren't frozen out, and tight harmonic trapping potential of frequency $\omega$ in the frozen dimensions.

@@

\ack
Acknowledgements in here; funding sources etc.  Joe's Future Fellowship.

@@

\appendix
\section{Derivation of two mode squeezing formula}
\label{appendixTwoModeDerivation}
In this Appendix we derive the equations for the number variances. We use the model described in Section \ref{secTwoModeAnalytic}, where we consider a two mode system with $N$ particles, initially prepared as a coherent state with all the population in mode $|a\rangle$ at time $t=0$. A coupling pulse is then applied, so that the system evolves under the Hamiltonian
\begin{equation}
\hat{H_1} = \hbar \omega \hat{a}^{\dagger} \hat{a} +  \hbar \omega \hat{b}^{\dagger} \hat{b} 
          + \Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger} \hat{a} )
\end{equation}
until time $t=t_1$. From time $t_1$ until time $t=t_2$ the coupling is turned off, and the system evolves under the Hamiltonian
\begin{equation}
\hat{H_2} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}.
\end{equation}
Finally, from time $t_2$ until time $t=t_3$, the coupling is once again applied with a phase offset of $\phi$ relative to the first time, and the system evolves via
\begin{equation}
\hat{H_3} = \hbar \omega \hat{a}^{\dagger} \hat{a} +  \hbar \omega \hat{b}^{\dagger} \hat{b}
          + \Omega (e^{i\phi} \hat{a}^{\dagger} \hat{b} + e^{-i\phi} \hat{b}^{\dagger} \hat{a} ).
\end{equation}
It is clear that the linear self-energy terms proportional to $\omega$ result in a constant phase rotation of the fields $\hat{a}$ and $\hat{b}$. As this system is self-contained and global phase is unmeasurable, all phase measurements can only rely on the difference in phase between the two modes. This phase rotation is therefore irrelevent and will be ignored in the following analysis.
